Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} \Large{Stagnating State Funding for Higher Education and its Effect on Faculty at Public Universities}

### 2023-12-19 text/sections/03-concepts.tex
Wiritng updates, new title.

While state support for public higher education has stagnated at the same time as education costs rose, there are multiple possible ways that a university can respond.
\cite{NBERw23736} established that corresponding rises in tuition did not offset the falling state support, so that university spending (per student) fell in response to these persistent, negative state funding shocks.
There are multiple ways that these changes in finances may affect the faculty composition at public universities.

It is a priori unclear that stagnating state support and changes in faculty composition at public universities would be causally linked.
Over the 21st Century there have been drastic changes to college selectivity, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, so that over time public universities (on average) substitute towards contingent lecturers. 
This is one way in which the trends may be concurrent, but not causal.

Employment composition changes likely arise via differences in hiring rate across levels of professor seniority.
Universities hire new professors most years, either to expand their departments or to replace leaving professors, but do not hire at the same rate for each level of seniority.
The yearly hiring is usually highest among non-tenured adjunct faculty, where universities hire instructors often on short-term or contingent contracts, so that lecturers are less costly to hire (or to let go) in response to yearly changes.
Tenure-track assistant professors are mostly hired with a four to six year contract, and formal agreement for tenure consideration at the end of this term.
Lastly, tenured professors have successfully secured a full-time appointment at their university with no expiration date; this position has the lowest yearly hiring rate among most universities.
Similarly, there is highest turn-over (including leaving the university) among lecturers, and lowest among full professors.
The differences in yearly hiring and turnover rate mean that if hiring is affected by state funding, then there will be faculty composition change in the years following budget cuts.

Faculty salaries may be affected by changes in public university finances.
When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires may be less likely to accept the lower salary offers from public universities.\footnote{
    Faculty often consider their outside options, as in offers for employment and/or promotion from other universities.
    See \cite{blackaby2005} for an overview of faculty outside options.
}
Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
    Indeed, Cornell University implemented hiring freezes and a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the cut was returned to professors in the year 2021.
}
Additionally, it is not immediately clear which faculty (among those already hired) will be most affected by the changes in their university's funding.
Yet, there is one empirical fact worth noting: the average assistant or tenured professor earns more than double that of a lecturer (see \autoref{tab:illinois-summary}).
If a public university's primary obligation is to teach, and they must fulfil this objective with less resources, then they may substitute away from tenure-track and tenured professors towards lecturers.

Similarly, its is not clear when the effects of funding short-falls will be realised.
University departments often employ faculty on multiple-year contracts (e.g., the tenure contract), and individual professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for university funding to change the faculty composition, or trickle down to faculty.
For example, university departments coordinate hiring in annual cycles, so that a budget short-fall in year $t$ may have no effect on the in-progress hiring committee and resulting hiring decisions, yet the hiring cycle in year $t+1$ may be postponed or cancelled in response.
As such, this paper investigates the dynamics of state funding shocks' effects on faculty, for the years following a university funding shock.

>>>>

Universities can respond multiple ways to the falls in per student state funding.
\cite{NBERw23736} established that public universities who experienced state funding cuts did not fully off-set by raising tution, and cut total spending.
But it is not clear how these spending cuts went on to affect faculty --- or whether they did at all.

Wider changes in the American higher education sector may concurrently explain the trends in state funding and faculty substitition.
College selectivity changed drastically over the 21st century, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, leaving public universities substituting toward lecturers over time.
This is one way in which the trends may be concurrent, but not causal.
I address this issue by using ssift-share instrument approach to isolate funding cuts that affected more funding-intensive universities, measuring the rate of substitution between lecturers and professors in response to yearly changes in state-wide funding.

Universities could respond to funding cuts by cutting faculty salaries, leading to high-performing faculty leaving the university.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.}
Funding cuts could also limit the amount public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting these offers.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time (see \citealt{blackaby2005} for an overview of faculty outside options).
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: the average assistant or tenured professor earns more than double that of a lecturer (see \autoref{tab:illinois-summary}).
Secondly, lecturers are (by definition) granted shorter contracts than professors; assistant professors mostly have 5--6 year long contracts, full professors indefinite contracts (tenure), lecturers yearly or term-wise contracts.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer resources, then substitution away from tenure-track and tenured professors towards lecturers is rationalisable.

It is not clear when the effects of funding short-falls will be realised.
University departments often employ faculty on multiple-year contracts (e.g., the tenure contract), and individual professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for university funding to change the faculty composition, or trickle down to faculty.
For example, university departments coordinate hiring in annual cycles, so that a budget short-fall in year $t$ may have no effect on the in-progress hiring committee and resulting hiring decisions, yet the hiring cycle in year $t+1$ may be postponed or cancelled in response.
As such, this paper investigates the dynamics of state funding shocks' effects on faculty, for the years following a state funding cut.
